### 2010-07-05 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix Hutton's example
Instead of fiddling with two separate finite domains for each types,
we really define a notion of context, environment, and variable
pointing into that.

Addressing the question of a reviewer, the model contains the case of
pairs, mixing any odd type.

A consequence is that we talk about equality. But that should be ok.

%% Types
\newcommand{\Ty}{\C{Ty}}
\newcommand{\Ebool}{\etag{\CN{bool}}}
\newcommand{\Enat}{\etag{\CN{nat}}}

@@

%% Index mapper (terminology?)
\newcommand{\SYMBVal}{\F{Val}\xspace}
\newcommand{\Val}[1]{\SYMBVal\:#1}
\newcommand{\SYMBVar}{\F{Var}\xspace}
\newcommand{\Var}[2]{\SYMBVar_{#1} #2}

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\SYMBVal$, it is possible to
build and manipulate \emph{open} terms, that is, terms with
variables. Following $\SYMBVal$, we define $\SYMBVar$ by:
%
\[\stk{
\Var{}{} : \EnumU \To \Ty \To \Set \\
\Var{\V{dom}}{\_} = \EnumT{\V{dom}}
}\]
%
Whereas $\SYMBVal$ was mapping the type to the corresponding host type,
$\SYMBVar$ maps types to a finite set. The finite set -- the context
-- contains closed terms. A variable is therefore a $\SYMBEval$ that
contains a pointer to a particular element of the finite set -- an
element of $\EnumT{\V{dom}}$.

Consequently, replacing $\Val{\V{ty}}$ by
$(\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})$ in
Figure~\ref{fig:hexpr-full} turns the language of closed terms into a
language of opened terms with variables and constants. For
readability, we abbreviate $\LAM{\V{ty}}
\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ by
$\SUM{\SYMBVal}{\Var{\V{dom}}{}}$. \note{Is it polite and
  comprehensible to ask that?} This defines a new indexed description,
called $\HExprVarD{\V{dom}}$.

@@

The $\discharge$ function is relative to the required type and a
context of the right type. Its action is to map values to themself,
and variables to their value in context. This corresponds to the
following function:
%
\[\stk{
\begin{array}{@{}ll}
\discharge : & \PITEL{\V{ty}}{\Ty}
               \PITEL{\V{dom}}{\EnumU} \\
             & \PI{\V{\gamma}}{\spi{\V{dom}}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})}} \\
             & (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To
               \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}} 
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumLeft \V{x})  & \Con{\pair{\SYMBEval}{\V{x}}{}} \\
\discharge\: \V{ty}\: \V{dom}\: \V{\gamma}\: (\SumRight \V{v}) &
\end{array}\\
\qquad\switch{\V{dom}}{(\LAM{\_}
\IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})}{\V{\gamma}}{\V{v}}
}\]

@@

\newcommand{\substH}{\F{substExpr}}
\newcommand{\domNat}{dom_{\CN{nat}}}
\newcommand{\domBool}{dom_{\CN{bool}}}
\newcommand{\gammaNat}{\V{\ensuremath{\gamma_{\CN{nat}}}}}
\newcommand{\gammaBool}{\V{\ensuremath{\gamma_{\CN{bool}}}}}
\newcommand{\SYMBGammaSpi}{\F{\ensuremath{\Gamma_{\CN{ty}}}}\xspace}
\newcommand{\GammaSpi}[2]{\SYMBGammaSpi\: #1\: #2}

We are now left with applying $\discharge$ over all variables of the
term. The type of this operation is the following:
%
\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{dom}}{\EnumU} \\
           & \PITEL{\V{\gammaNat}}{\GammaSpi{\V{dom}}{\Enat}} 
             \PITEL{\V{\gammaBool}}{\GammaSpi{\V{dom}}{\Ebool}} \\
           & \begin{array}{@{}ll}
             \PI{\V{\sigma}}{& \PITEL{\V{dom}}{\EnumU} \\
                             & \PITEL{\V{\gammaNat}}
                                     {\GammaSpi{\V{dom}}{\Enat}} \\
                             & \PITEL{\V{\gammaBool}}
                                     {\GammaSpi{\V{dom}}{\Ebool}} \\
                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To \\
                             & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}}
             \end{array}\\
          & \PI{\V{ty}}{\Ty}
            \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}} \To \\
          & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
\end{array}
\]
%
Where $\SYMBGammaSpi$ corresponds to a context, defined by:
%
\[\stk{
\SYMBGammaSpi : \EnumU \To \Ty \To \Set                                             \\
\GammaSpi{\V{dom}}{\V{ty}} = \spi{\V{dom}}{(\LAM{\_} \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}})} 
}\]

Abstracting away the book-keeping introduced by contexts, this
definition looks familiar. It is similar to a monadic \bind. This is
not surprising as we are defining a first-order syntax with variables:
our datatype enjoys more structure than what we are given. We are
facing a free monad, where $\SYMBEval$ is the \return introducing
variables. For convenience, we wrap $\discharge$ in a $\V{\sigma}$
function that picks the context of the right type:
%
\[
\F{$\sigma$}\: \V{dom}\: \gammaNat\: \gammaBool\: \V{ty}\: \V{var} \mapsto
    \F{$\discharge$}\: \V{ty}\: \V{dom}\: \F{$\gamma$}_{\V{ty}}\: \V{var} 
\]
%
Where $\F{$\gamma$}_{\V{ty}}$ is short for
 $\caseB{\V{ty}}{\Enat  \To \gammaNat \\
                 \Ebool \To \gammaBool}$

Instead of implementing $\substH$ in this special case, we are now
going to implement the \emph{free indexed-monad construction}.

@@

In Section~\ref{sec:idesc-examples}, we had the intuition that our
datatypes $\HExprD$ and $\HExprVarD{\V{dom}}$ enjoy a monadic
structure. We had identified the variable substitution operation as
the \bind of a free monad. To exhibit its monadic structure, we first
have to massage the definition of our datatype.

@@

We call this datatype $\HExprFreeD$. By a simple unfolding of
definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}$
corresponds to the syntax of closed terms, $\HExprD$. Similarly,
$\FreeIMonad{\HExprFreeD}{\Ty}{(\SUM{\SYMBVal}{\Var{\V{dom}}{}})}$
corresponds to expressions with variables, $\HExprVarD{\V{dom}}$.

@@

The evaluator for closed terms we implemented in
Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}\: \V{ty}$ to values in
$\Val{\V{ty}}$. We are left with implementing $\substH$. We simply
have to fill in the right arguments to $\substI$, the type guiding us:
%
\[\stk{
\substH\: \V{dom}\:
          \gammaNat\: \gammaBool\:
          \V{\sigma}\: 
          \V{ty}\: 
          \V{term} \mapsto  \\
\;\;\ \begin{array}{ll}
       \substI_{\Ty}\: &  \HExprFreeD\: 
                         (\SUM{\SYMBVal}{\Var{\V{dom}}{}})\: 
                         \SYMBVal \\
                      &
                        (\V{\sigma}\: \V{dom}\: \gammaNat\: \gammaBool)\:
                        \V{ty}\:
                        \V{term}
      \end{array}
}\]
%
Hence completing our implementation of the open terms interpreter.

>>>>

%% Types
\newcommand{\Ty}{\D{Ty}}
\newcommand{\Ebool}{\etag{\CN{bool}}}
\newcommand{\Enat}{\etag{\CN{nat}}}

@@

%% Index mapper (terminology?)
\newcommand{\SYMBVal}{\F{Val}\xspace}
\newcommand{\Val}[1]{\SYMBVal\:#1}
\newcommand{\SYMBVar}{\F{Var}\xspace}
\newcommand{\Var}[2]{\SYMBVar\: #1\: #2}

@@

\newcommand{\Ctxt}{\D{Context}}
\newcommand{\SYMBCtxtEmpty}{\C{[]}\xspace}
\newcommand{\CtxtEmpty}{\SYMBCtxtEmpty}
\newcommand{\SYMBCtxtSnoc}{\C{snoc}\xspace}
\newcommand{\CtxtSnoc}[2]{\SYMBCtxtSnoc\:#1\:#2}
\newcommand{\SYMBEnv}{\F{Env}}
\newcommand{\Env}[1]{\SYMBEnv\: #1}
\newcommand{\SYMBlookup}{\F{lookup}}
\newcommand{\lookup}[4]{\SYMBlookup\: #1\: #2\: #3\: #4}

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\SYMBVal$, it is possible to
build and manipulate \emph{open} terms, that is, terms in a
context. The context is represented by a snoc-list of types:
%
\[
\stk{
\data \Ctxt : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \SYMBCtxtEmpty  & \Ctxt   \\
    \SYMBCtxtSnoc   & \Ctxt \To \Ty \To \Ctxt
\end{array}
}\]
%
An environment realises the context, associating a value for each type:
%
\[
\stk{
\SYMBEnv : \Ctxt \To \Set \\
\begin{array}{@{}l@{\:\:=\:\:}l}
\Env{\CtxtEmpty}        & \Unit \\
\Env{(\CtxtSnoc{G}{S})} & \TIMES{\Env{G}}{\Val{S}}
\end{array}
}\]
%
In this setting, we define $\SYMBVar$ by:
%
\[\stk{
\Var{}{} : \Ctxt \To \Ty \To \Set \\
\begin{array}{@{}ll@{\:\:=\:\:}l}
\Var{\CtxtEmpty}{& T}            & \Void \\
\Var{(\CtxtSnoc{G}{S})}{& T}     & \SUM{(\Var{G}{T})}{(S \Eq T)}
\end{array}
}\]
%
While $\SYMBVal$ was mapping the type to the corresponding host type,
$\SYMBVar$ indexes a value in the context, obtaining a proof that the
types matches. The $\SYMBlookup$ function precisely follow this
semantics:
%
\[\stk{
\SYMBlookup : \PI{\V{G}}{\Ctxt} 
          \Env{\V{G}} \To 
          \PI{\V{T}}{\Ty} 
          \Var{\V{G}}{\V{T}} \To
          \Val{\V{T}} \\
\begin{array}{@{}lll}
\lookup{\CtxtEmpty}{               \_}{                   \V{T}}{()}                    &   & \\
\lookup{(\CtxtSnoc{\V{G}}{.T})}{   \pair{\V{g}}{\V{t}}{}}{\V{T}}{(\SumRight{\C{refl}})} & = & \V{t} \\
\lookup{(\CtxtSnoc{\V{G}}{\V{S}})}{\pair{\V{g}}{\V{t}}{}}{\V{T}}{(\SumLeft{\V{x}})}     & = & \lookup{\V{G}}{\V{g}}{\V{T}}{\V{x}} 
\end{array}
}\]

Consequently, replacing $\Val{\V{ty}}$ by
$(\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}})$ in
Figure~\ref{fig:hexpr-full} turns the language of closed terms into a
language of opened terms with variables and constants. For
readability, we abbreviate $\LAM{\V{ty}}
\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}$ by
$\SUM{\SYMBVal}{\Var{\V{G}}{}}$. \note{Is it polite and
  comprehensible to ask that?} This defines a new indexed description,
called $\HExprVarD{\V{G}}$.

@@

The $\discharge$ function is relative to the required type and a
context of the right type. Its action is to map values to themself,
and variables to their value in context. This corresponds to the
following function:
%
\[\stk{
\begin{array}{@{}ll}
\discharge : & \PI{\V{G}}{\Ctxt}
               \Env{G} \To
               \PITEL{\V{ty}}{\Ty} \\
             & (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
               \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} 
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumLeft \V{x})  & 
    \Con{\pair{\SYMBEval}{\V{x}}{}} \\
\discharge\: \V{G}\: \V{g}\: \V{ty}\: (\SumRight \V{v}) &
    \Con{\pair{\SYMBEval}{\lookup{\V{G}}{\V{g}}{\V{ty}}{\V{v}}}{}}
\end{array}
}\]

@@

\newcommand{\substH}{\F{substExpr}}

We are now left with applying $\discharge$ over all variables of the
term. The type of this operation is the following:
%
\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{ty}}{\Ty}
             \PI{\V{G}}{\Ctxt} \\
           & \PITEL{\V{\sigma}}{\PI{\V{ty}}{\Ty}
                                (\SUM{\Val{\V{ty}}}{\Var{\V{G}}{\V{ty}}}) \To
                                \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}} \\
          & \To \IMu{\Ty}{\toIDesc{\HExprVarD{\V{G}}}}{\V{ty}} \To 
            \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
\end{array}
\]

Abstracting away the book-keeping introduced by contexts, this
definition looks familiar. It is similar to a monadic \bind. This is
not surprising as we are defining a first-order syntax with variables:
our datatype enjoys more structure than what we are given. We are
facing a free monad, where $\SYMBEval$ is the \return introducing
variables. Instead of implementing $\substH$ in this special case, we
are now going to implement the \emph{free indexed-monad construction}.

@@

In Section~\ref{sec:idesc-examples}, we had the intuition that our
datatypes $\HExprD$ and $\HExprVarD{\V{G}}$ enjoy a monadic
structure. We had identified the variable substitution operation as
the \bind of a free monad. To exhibit its monadic structure, we first
have to massage the definition of our datatype.

@@

We call this datatype $\HExprFreeD$. By a simple unfolding of
definition, we note that $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}$
corresponds to the syntax of closed terms, $\HExprD$. Similarly,
$\FreeIMonad{\HExprFreeD}{\Ty}{(\SUM{\SYMBVal}{\Var{\V{G}}{}})}$
corresponds to expressions with variables, $\HExprVarD{\V{G}}$.

@@

The evaluator for closed terms we implemented in
Section~\ref{sec:idesc-examples} remains unchanged. It reduces closed
terms in $\FreeIMonad{\HExprFreeD}{\Ty}{\SYMBVal}\: \V{ty}$ to values in
$\Val{\V{ty}}$. We are left with implementing $\substH$. We simply
have to fill in the right arguments to $\substI$, the type guiding us:
%
\[\stk{
\substH\: \V{ty}\:
          \V{G}\:
          \V{g}\:
          \V{\sigma}\: 
          \V{term} \mapsto  \\
\quad \substI_{\Ty}\:
               (\SUM{\SYMBVal}{\SYMBVar\: \V{G}})\: 
               \SYMBVal\:
               \HExprFreeD\: 
               \V{\sigma}\:
               \V{ty}\:
               \V{term}
}\]
%
Hence completing our implementation of the open terms interpreter.
